{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-07-25 Docs/manuscript.tex
Updated discussion

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}. % emwJun4 -- you need to use ` to open single quotes in Latex. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (number of studies = 12). See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.}  % emwJun4 -- add to caption how many studies in europe (7?)
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.

@@

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used Bayesian hierarchical models. We regressed DOY of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in rstanarm package (version 2.21.3; \citealp{brilleman2018}) using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples. We checked for model fit by ... [ADD here]. We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated. 
%emw2Mar -- It would be good if you have tidy code you can post with the data so folks can fit the models themselves
% emwJun4 -- do you need help with ADD here ... if so, ping me in the old issue and we can work on it. It should be easy if you have cleared up all divergent transitions!
% We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.

@@

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events were earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

@@

% emwJun4 -- can you reference the figures in the supp that have the latitude results?

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms (). Spring events weakly diverged, delaying at a rate of 0.37 (0.15 - 0.59) days per degree north in angiosperms and advancing 0.23 (0.00 - 0.46) days per degree north in gymnosperms. Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
Effects of MAT on spring events also weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig. \ref{figure:continental_spptype_effect}B).

% emwJun4 -- give more explanation of what we mean by  'in months that matter for the events' ... is this in the methods? 
While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig. \ref{supplement-figure:overlap_scatterplot} in Supplements). We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one per cent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one per cent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).

@@

Opening (no subheader) 
\begin{enumerate}
\item Overall, our results show inconsistent and weak clines in spring events across continents, while fall events generally showed much stronger clines, especially in North America
\begin{enumerate}
\item For spring phenology there were weak diverging effects by species groupings -- which may be important, and suggest different trends that mute an overall response. 
\item Our results suggest that spring events are highly plastic, and thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures.  %7July2023 -- careful when talking about plasticity and adaptation; our results can be suggestive but do not test either of these things directly
\end{enumerate}
\item  These interesting trends by species grouping for spring phenology and continent for fall phenology, suggests data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, would aid forecasting. 
\begin{enumerate}
\item We found MAT and latitude still good predictors, better than climate overlap or other metrics. 
\item While coarse metrics such as latitude and MAT appear to represent how similar the climates are between the provenances and gardens in times that matter for the events, at least in the narrow range of northern temperature and boreal forests we studied
\end{enumerate}
\end{enumerate}
%7July2023 -- I don't understand the below: where do we show these continent effects for spring phenology?
% The weak relationship between spring event dates and provenance latitude and MAT that we find in European studies might be explained by the higher extent of climate overlap in those studies. The more similar the climate is between provenances and gardens, the less difference between spring event dates.

Subheader: Variation across continents and species groupings
\begin{enumerate}
\item Species differences
\begin{enumerate}
\item Angiosperm vs gymnosperm showed some effects, which makes sense as timing of leafout here is very different (gymnosperem photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost)
\item Suggests there may be additional differences if we have more species and more information on leafout timing and frost risk approach ... 
\item discuss xylem (ring versus diffuse porous spp) and embolism  (see Salk article)
\item Maybe also here: Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}. 
\end{enumerate}
\item Continental differences
\begin{enumerate}
\item No strong evidence of continental differences, could make sense based on long-term climate history perhaps ... 
\item Maybe also discuss ice sheet dynamics in Europe/North America, which could impact rates of adaptation.
\end{enumerate}
\end{enumerate}

Subheader: Adaptation in spring vs fall phenology: Implications for climate change responses
\begin{enumerate}
\item Our results suggest that current advances in spring phenology may continue, but delays in fall phenology could depend on speed of adaptation
\begin{enumerate}
\item Spring events appear highly plastic from these studies, which are good tests. 
\item Some previous chamber studies finding potential adaptation (cited in Ettinger et al. 2020 Nature Climate Change, I believe -- check the paper and supp) but these could be influenced by chilling at the provenance site, whereas gardens do not have this effect
\item Previous work has suggested strong clines in fall phenology and our findings back this up... 
\item These findings have important implications for growing discussion of how shifting phenology and tree growth relate as our results suggest there are strong limits on shifts in fall phenology (Zohner2023 and other papers such as Zani 2020 -- need a couple sentences here) % D. Zani, T. W. Crowther, L. Mo, S. S. Renner, C. M. Zohner, Increased growing-season productivity drives earlier autumn leaf senescence in temperate trees. Science 370, 1066–1071 (2020).
\end{enumerate}
\item These trends likely vary by species and depend on frost tolerance strategy
\begin{enumerate}
\item If current advances in spring phenology continues it could mean shifts in spring frost etc., but differences across species suggest this will not impact all species the same. 
\item In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19} ... further Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}....  and species vary in how the balance this tradeoff. 
\end{enumerate}
\item Species may also vary in whether we predict co- or counter-gradient relationships with spring phenology
\begin{enumerate}
\item Fall events were always co-gradient
\item Spring events were more varied -- some appeared to be co or counter ... 
\item Add some discussion of what this might mean (one idea is that spring leafout sets up flowering and thus some counter-gradient could aid gene flow over larger areas? Other ideas you have?) and cite some of the relevant papers. 
\end{enumerate}
\item All of this critical to predicting not just phenological change, but range expansions with climate change: An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
\begin{enumerate}
\item Implications of small differences in spring phenology could mean different gene flow etc. 
\item the degree of adaptation dictated by climate change also differs across different geographical areas \citep{Loarie09}. 
\item Mention elevational results here? 
\end{enumerate}
\end{enumerate}

>>>>

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}. % emwJun4 -- you need to use ` to open single quotes in Latex. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.\ref{supplement-table:all_studies} in Supplements for information on selected studies.} 
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.

@@

\subsection{Analyses}
To estimate clines in spring and fall phenological events across species we used Bayesian hierarchical models. We regressed DOY of events against geographical and climate predictors with partial pooling (sometimes called `random effects') on the intercept and slope for each species within each garden. Because most tree species were present in only one common garden in our dataset, it was impossible to fit garden and species separately, thus we treat each species within a garden as a unique group. Using posterior estimates for each species within a garden, we estimated effects of continent (North America vs. Europe) and species type (angiosperm vs. gymnosperm). All models were fit in `rstanarm' package (version 2.21.3; \citealp{brilleman2018}) using default priors, with 4 chains and 1000 sampling iterations per chain for a total of 4000 samples. We checked for model fit by ... [ADD here]. We present estimates as mean $\pm$ 90\% uncertainty intervals given parenthetically, unless otherwise stated.

% alinaJuly20 I need help with ADD here (ping'd you in the old issue. I have removed all divergent transitions already)

% We pooled the timing of budburst and leaf flush into a single category of ‘spring events’ and the timing of bud set, leaf senescence, growth cessation, and leaf abscission into ‘fall events.’ Such pooling is justified because of the shared pressures from natural selection that govern these events \citep{Gill15}.

@@

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events were earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe such relationship was weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

@@

Effects of provenance latitude on fall events were similar across angiosperms and gymnosperms (Fig.\ref{figure:continental_spptype_lat}B). Spring events weakly diverged, delaying at a rate of 0.37 (0.15 - 0.59) days per degree north in angiosperms and advancing 0.23 (0.00 - 0.46) days per degree north in gymnosperms. Fall events advanced 3.18 (2.76 - 3.62) days per degree north in angiosperms and 3.14 (2.81-3.47) days per degree north in gymnosperms.
Effects of MAT on spring events also weakly diverged (Fig.\ref{figure:continental_spptype_effect}B). Spring events advanced 0.82 (0.54 - 1.11) days/$^{\circ}$C as MAT increased in angiosperms and delayed 0.76 (0.37 - 1.14) days/$^{\circ}$C as MAT increased in gymnosperms. Fall events delayed in warmer locations for both species types, but slightly more so for gymnosperms (6.23 days) than angiosperms (3.69 days) (Fig. \ref{figure:continental_spptype_effect}B).

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig. \ref{supplement-figure:overlap_scatterplot} in Supplements). We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one percent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one percent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).

@@

Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe. Specifically, there were weak diverging effects by species x garden groupings, which might be important and suggest different trends that mute an overall response. Fall events generally showed much stronger clines, suggesting a higher degree of local adaptation, especially in North America. While not directly testing plasticity or adaptation, our study suggests that spring events are highly plastic, and thus we predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures. These interesting trends by species grouping for spring events and continent for fall events suggest that data on more species and greater information on important factors, such as their geographic location in relation to their origins and elevation, would aid forecasting. 
We found coarse metrics such as provenance latitude and MAT to be better predictors of timing than climate overlap and other metrics. Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in months close to the events, at least in the narrow range of northern temperature and boreal forests we studied.

\subsection{Variation across continents and species groupings}

Our results indicated that the effects of provenance latitude and MAT on fall events were similar across angiosperm and gymnosperm species: Fall events delay as MAT increases for both leaf types, slightly more so for gymnosperms, explained by sensitivity to day-length signals. However, the two leaf types exhibited slightly diverging trends in spring events:  advancing in angiosperms and delaying in most gymnosperms as MAT increases. Such differences make sense as timing of budburst and leafout varies by leaf type. Gymnosperms photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost \citep{panchen14}. Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}. 
Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

\begin{enumerate}
\item Maybe also discuss ice sheet dynamics in Europe/North America, which could impact rates of adaptation.
\end{enumerate}
While we noticed more prominent clines in fall events in North America, there was no strong evidence of overall continental differences in spring events, which might again suggest a higher degree of plasticity in spring events. Meanwhile, most North American gardens and provenances included in this study were along the Pacific Northwest region \ref{}, which might mask the kind of climate pattern we were hoping to see from the entirety of North America.  % alinaJuly20: not sure about how to link to long-term climate history.

\subsection{Adaptation in spring vs fall phenology: Implications for climate change responses}
\begin{enumerate}
\item Our results suggest that current advances in spring phenology may continue, but delays in fall phenology could depend on speed of adaptation
\begin{enumerate}
\item Spring events appear highly plastic from these studies, which are good tests. 
\item Some previous chamber studies finding potential adaptation (cited in Ettinger et al. 2020 Nature Climate Change, I believe -- check the paper and supp) but these could be influenced by chilling at the provenance site, whereas gardens do not have this effect
\item Previous work has suggested strong clines in fall phenology and our findings back this up... 
\item These findings have important implications for growing discussion of how shifting phenology and tree growth relate as our results suggest there are strong limits on shifts in fall phenology (Zohner2023 and other papers such as Zani 2020 -- need a couple sentences here) % D. Zani, T. W. Crowther, L. Mo, S. S. Renner, C. M. Zohner, Increased growing-season productivity drives earlier autumn leaf senescence in temperate trees. Science 370, 1066–1071 (2020).
\end{enumerate}
\item These trends likely vary by species and depend on frost tolerance strategy
\begin{enumerate}
\item If current advances in spring phenology continues it could mean shifts in spring frost etc., but differences across species suggest this will not impact all species the same. 
\item In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19} ... further Plants that budburst earlier can take advantage of a longer growing season and avoid competition  \citep{guo22}....  and species vary in how the balance this tradeoff. 
\end{enumerate}
\item Species may also vary in whether we predict co- or counter-gradient relationships with spring phenology
\begin{enumerate}
\item Fall events were always co-gradient
\item Spring events were more varied -- some appeared to be co or counter ... 
\item Add some discussion of what this might mean (one idea is that spring leafout sets up flowering and thus some counter-gradient could aid gene flow over larger areas? Other ideas you have?) and cite some of the relevant papers. 
\end{enumerate}
\item All of this critical to predicting not just phenological change, but range expansions with climate change: An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction.
\begin{enumerate}
\item Implications of small differences in spring phenology could mean different gene flow etc. 
\item the degree of adaptation dictated by climate change also differs across different geographical areas \citep{Loarie09}. 
\item Mention elevational results here? 
\end{enumerate}
\end{enumerate}
